### 2023-05-18 proofs/main.tex
Add soundness of solution merge.

\renewcommand{\ottdruleSMESupSupName}[0]{$([[≥]]\&[[≥]])$}
\renewcommand{\ottdruleSMEEqSupName}[0]{$([[≈]]\&[[≥]])$}
\renewcommand{\ottdruleSMESupEqName}[0]{$([[≥]]\&[[≈]])$}
\renewcommand{\ottdruleSMEPEqEqName}[0]{$([[≈]]\&[[≈]]^{+})$}
\renewcommand{\ottdruleSMENEqEqName}[0]{$([[≈]]\&[[≈]]^{-})$}

@@

@@

@@

Two matching entries can be merged in the following way:
\begin{definition} \hfill \\
\ottdefnSME\\
\end{definition}
Notice that in case of equivalence, the assigned types
must be equal (i.e. alpha-equivalent) to be merged. This is because
the unification algorithm assumes that every type is normalized,
and hence, equivalence is alpha-equivalence 
(\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).

>>>>

@@

% Γ ⊢ usEntry1 & usEntry2 = usEntry3
\renewcommand{\ottdruleSMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
\renewcommand{\ottdruleSMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
\renewcommand{\ottdruleSMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
\renewcommand{\ottdruleSMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
\renewcommand{\ottdruleSMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}

% Γ ⊢ usEntry1 ⇒ usEntry2
\renewcommand{\ottdruleSImpESupSupName}[0]{$([[≥]]\Rightarrow^{+}[[≥]])$}
\renewcommand{\ottdruleSImpEEqSupName}[0]{$([[≈]]\Rightarrow^{+}[[≥]])$}
\renewcommand{\ottdruleSImpEPEqEqName}[0]{$([[≈]]\Rightarrow^{+}[[≈]])$}
\renewcommand{\ottdruleSImpENEqEqName}[0]{$([[≈]]\Rightarrow^{-}[[≈]])$}

@@

\subsection{Unification Solution Weakening}

@@

Unification solutions are preordered by the weakening relation.
First, let us define the weakening on the unification solution entries.\\

\begin{definition} \hfill \\
\ottdefnSImpE{}\\
\end{definition}
Notice that  $[[Γ ⊢ usEntry1 ⇒ usEntry2]]$ means that $[[usEntry1]]$ and $[[usEntry2]]$ are matching. And 
matching is a equivalence relation on the set of unification solutions.

Next, we lift the weakening relation to unification solutions.
Informally, $[[Θ ⊢ us1 ⇒ us2]]$ means that for any entry of $[[us2]]$,
there is a matching entry in $[[us1]]$ that is stronger than it.

\begin{definition}
  Assuming $[[us2 : Θ]]$ and $[[us1 : Θ']] \supseteq [[Θ]]$,
  we say $[[Θ ⊢ us1 ⇒ us2]]$ iff
  $\forall [[usEntry2]] \in [[us2]]$, $\exists [[usEntry1]] \in [[us1]]$ s.t.
  \begin{enumerate}
    \item $[[usEntry1]]$ and $[[usEntry2]]$ are matching, i.e. restricting the same unification variable $[[α̂±]]$;
    \item $[[Θ(α̂±) ⊢ usEntry1 ⇒ usEntry2]]$ (where $[[Θ(α̂±)]]$ is the context corresponding to $[[α̂±]]$ in $[[Θ]]$).
  \end{enumerate}
\end{definition}

Two matching entries can be merged in the following way:
\begin{definition} \hfill \\
\ottdefnSME\\
\end{definition}
% Notice that in case of equivalence, the assigned types
% must be equal (i.e. alpha-equivalent) to be merged. This is because
% the unification algorithm assumes that every type is normalized,
% and hence, equivalence is alpha-equivalence 
% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).

### 2023-05-18 proofs/tex/alg-subtyping-lemmas.tex
Add soundness of solution merge.

\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
    \begin{itemize}
        \item [$-$] 
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
        implies $[[us : Θ|uv uN]]$ and 
        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
        $[[ Γ ⊢ [us']uN ≤ iM ]]$

        \item [$+$] 
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
        implies $[[us : Θ|uv uP]]$ and 
        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
        $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
     \end{itemize}
\end{theorem}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ (and mutually, on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$).
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
        Then $[[uv a⁻ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
        Since $[[uv a⁻ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁻]]$, i.e.
        $[[ [us']a⁻ = a⁻ ]]$. Therefore, $[[Γ ⊢ [us']a⁻ ≤ a⁻]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.

@@

        \item $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
        The next step of the algorithm is two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
        By the induction hypothesis, 
        \begin{enumerate}
            \item $[[us1 : Θ | uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1']]$ s.t. $[[Θ ⊢ us1' ⇒ us1]]$
            \item $[[us2 : Θ | uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2']]$ s.t. $[[Θ ⊢ us2' ⇒ us2]]$
        \end{enumerate}

@@

        Notice that $[[Γ, pbs ⊢ Θ, â⁺*[G, pbs] ]]$, $[[Γ,pbs ⊢ iM']]$, and 
        $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN ]]$, so the induction hypothesis is applicable,
        that is $[[us' : Θ | uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any
        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.

@@

    \item $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$\\

    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
    transitivity (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
    as required.

    \item For the other positive cases, the proof is symmetric to the corresponding negative cases.
    \end{caseof}
\end{proof}

Notice that the induction statement in following theorem is not symmetric for the positive and negative cases.
\begin{theorem}[Completeness and Initiality of the Subtyping Algorithm]
    \hfill
    \begin{itemize}
        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
        $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
        and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
        Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
        and $[[Θ ⊢ us ⇒ us']]$.

        \item [$+$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
        there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
        Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
        and $[[Θ ⊢ us ⇒ us']]$.
    \end{itemize}
\end{theorem}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$ (and mutually, on $[[ Γ ⊢ [us]uP ≥ iQ ]]$).
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$ or $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
    but first consider the base case for the substitution $[[ [us]uP ]]$:
    \begin{caseof}
        \item \label{case:subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$.
        $[[us : Θ | uv α̂⁺]]$ means that  $[[us]]$ is either $[[ (α̂⁺ :≈ iP') ]]$ or $[[ (α̂⁺ :≥ iP') ]]$,
        where $[[Δ ⊢ iP']]$ and $[[α̂⁺[Δ] ∊ Θ]]$.
        $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$
        because multiple inversions of \ruleref{\ottdruleDOneExistsLabel} 
        gives us $[[Γ ⊢ [us]α̂⁺ ≥ iQ]]$ since $[[ {nbs} ∩ fv [us]α̂⁺ = ∅]]$.

@@

        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel},
        \label{case:subt-complete-upshift}
        i.e. $[[ [us]uN ]] = [[ ↑[us]uP ]]$ (since the substitution $[[ [us]uN ]]$ must preserve the 
        top-level constructor of $[[uN]]\neq [[α̂⁻]]$), and $[[uM]] = [[ ↑iQ ]]$,
        and by inversion, $[[ Γ ⊢ [us]uP ≈ iQ ]]$.

@@

       It is left to show that $[[Θ ⊢ us ⇒ us'1 & us'2]]$.
       Since $[[us | uv uP ⊆ us]]$, 
       by \cref{lemma:weakening-monotonicity}, 
       $[[Θ ⊢ us ⇒ us | uv uP]]$,
       and then by \cref{lemma:weakening-transitivity}, 
       $[[Θ ⊢ us | uv uP ⇒ us'1]]$.
       Analogously, $[[Θ ⊢ us | uv uN' ⇒ us'2]]$.
       Then since by \cref{lemma:merge-completeness}, 
       $[[us'1 & us'2]]$ is the weakest restriction  
       implying both $[[us'1]]$ and $[[us'2]]$,
       we have $[[Θ ⊢ us ⇒ us'1 & us'2]]$, as required.

@@

       This way, the algorithm terminates with the result $[[us0' \ {â⁺*}]]$
       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁺*} ]]$. Notice that
       $[[us0' \ {â⁺*} : Θ | uv uN' ]]$. This way, to show $[[ Θ ⊢ us ⇒ us0' \
       {â⁺*} ]]$. it suffices to consider an unification variable $[[β̂±]] ∊
       [[uv uN']]$ and show that $[[Γ, pbs ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
       reflexivity of weakening (\cref{lemma:weakening-reflexivity}) since
       $[[us0'(β̂±)]] = [[us(β̂±)]]$.

      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
      \cref{case:subt-complete-nvar}.
      Notice that $[[ [us]uP = β⁺ ]]$ 
      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
      (\cref{case:subt-complete-base}).
    \end{caseof}
\end{proof}

>>>>

\begin{lemma}[Soundness of Positive Subtyping] \label{lemma:pos-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
    implies $[[us : Θ|uv uP]]$ and 
    for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
    $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
\end{lemma}
\begin{proof} 
    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$. Let us consider the last rule to infer this judgment.
    \begin{caseof}
    \item \ruleref{\ottdruleAPUVarLabel} then
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$.

    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
    transitively (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
    as required.

    \item  \ruleref{\ottdruleAPVarLabel} then
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
    Then $[[uv a⁺ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
    Since $[[uv a⁺ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁺]]$, i.e.
    $[[ [us'] a⁺ = a⁺]]$. Therefore, $[[Γ ⊢ [us']a⁺ ≥ a⁺]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.

    \item \label{case:pos-subt-soundness:arrow} 
     \ruleref{\ottdruleAShiftDLabel} then
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ↓uN ≥ ↓iM ⫤ us]]$.\\
    Then the next step of the algorithm is the unification of $[[nf(uN)]]$ and $[[nf(iM)]]$.
    By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
    it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uN]]$.
    By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 
    $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uN ]]$ as substitutions.
    $[[ [us]nf(uN) = nf(iM) ]]$ implies $[[Γ ⊢ [us]nf(uN) ≈ nf(iM)]]$, and then 
        $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$. \ilyam{add lemmas}

        Let us rewrite the left-hand side and the right-hand side of $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$ by 
        transitivity of equivalence (\cref{corollary:equivalence-transitivity}).
        By \cref{corollary:nf-sound-wrt-subt-equiv,corollary:subst-pres-equiv},
        $[[Γ ⊢ [us']nf(uN) ≈ [us']uN ]]$. By \cref{corollary:nf-sound-wrt-subt-equiv}, 
        $[[Γ ⊢ nf(iM) ≈ iM ]]$. 
        This way, we have $[[Γ ⊢ [us']uN ≈ iM]]$.
        Then by \ruleref{\ottdruleDOneShiftULabel}
        and congruence of substitution, $[[Γ ⊢ [us']↓uN ≥ ↓iM]]$.

    \item \label{case:pos-subt-soundness:exists}
       \ruleref{\ottdruleAExistsLabel} then
        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ∃nas.uP' ≥ ∃nbs.iQ' ⫤ us]]$ s.t. either $[[nas]]$ or $[[nbs]]$ is not empty.\\
        Then the algorithm creates fresh unification variables $[[â⁻*[Γ,nbs] ]]$, 
        substitutes the old $[[nas]]$ with them in $[[uP']]$, and makes the recursive call:
        $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ [â⁻*/nas] uP ≥ iQ ⫤ us']]$, returning as the result
        $[[us]] = [[us' \ {α̂⁻*}]]$.

        Notice that $[[Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$, $[[Γ,nbs ⊢ iQ']]$, and 
        $[[Γ,nbs; Θ, â⁻*[Γ, nbs] ⊢ [â⁻*/nas] uP' ]]$, so the induction hypothesis is applicable,
        that is $[[us' : Θ | uv [â⁻*/nas]uP']]$ and $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$ for any
        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.

    \end{caseof}
\end{proof}

\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
        implies $[[us : Θ|uv uN]]$ and 
        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
        $[[ Γ ⊢ [us']uN ≤ iM ]]$
\end{lemma}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$.
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
        Then $[[uv a⁻ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
        Since $[[uv a⁻ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁻]]$, i.e.
        $[[ [us']a⁻ = a⁻ ]]$. Therefore, $[[Γ ⊢ [us']a⁻ ≤ a⁻]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.

@@

        \item $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
        The next step of the algorithm is two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
        By the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}) and induction hypothesis, respectively 
        \begin{enumerate}
            \item $[[us1 : Θ | uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1']]$ s.t. $[[Θ ⊢ us1' ⇒ us1]]$
            \item $[[us2 : Θ | uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2']]$ s.t. $[[Θ ⊢ us2' ⇒ us2]]$
        \end{enumerate}

@@

        Notice that $[[Γ, pbs ⊢ Θ, â⁺*[G, pbs] ]]$, $[[Γ,pbs ⊢ iM']]$, and 
        $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN' ]]$, so the induction hypothesis is applicable,
        that is $[[us' : Θ | uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any
        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.

@@

\end{caseof}
\end{proof}

\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
\end{theorem}
\begin{proof}
\end{proof}

\begin{lemma}[Completeness of the Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
    there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
    and $[[Θ ⊢ us ⇒ us']]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
    Let us consider the last rule used in the derivation of  $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
    but first consider the base case for the substitution $[[ [us]uP ]]$:

    \begin{caseof}
        \item \label{case:subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$.
        $[[us : Θ | uv α̂⁺]]$ means that  $[[us]]$ is either $[[ (α̂⁺ :≈ iP') ]]$ or $[[ (α̂⁺ :≥ iP') ]]$,
        where $[[Δ ⊢ iP']]$ and $[[α̂⁺[Δ] ∊ Θ]]$.
        $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$
        because multiple inversions of \ruleref{\ottdruleDOneExistsLabel} 
        gives us $[[Γ ⊢ [us]α̂⁺ ≥ iQ]]$ since $[[ {nbs} ∩ fv [us]α̂⁺ = ∅]]$.

@@

      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
      \cref{case:subt-complete-nvar}.
      Notice that $[[ [us]uP = β⁺ ]]$ 
      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
      (\cref{case:subt-complete-base}).

    \end{caseof}
\end{proof}

\begin{lemma}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
    $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
    and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
    and $[[Θ ⊢ us ⇒ us']]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel},
        \label{case:subt-complete-upshift}
        i.e. $[[ [us]uN ]] = [[ ↑[us]uP ]]$ (since the substitution $[[ [us]uN ]]$ must preserve the 
        top-level constructor of $[[uN]]\neq [[α̂⁻]]$), and $[[uM]] = [[ ↑iQ ]]$,
        and by inversion, $[[ Γ ⊢ [us]uP ≈ iQ ]]$.

@@

       It is left to show that $[[Θ ⊢ us ⇒ us'1 & us'2]]$.
       Since $[[us | uv uP ⊆ us]]$, 
       by \cref{lemma:weakening-monotonicity}, 
       $[[Θ ⊢ us ⇒ us | uv uP]]$,
       and then by transitivity (\cref{lemma:weakening-transitivity}), 
       $[[Θ ⊢ us ⇒ us'1]]$.
       Analogously, $[[Θ ⊢ us ⇒ us'2]]$.
       Then since by \cref{lemma:merge-completeness}, 
       $[[us'1 & us'2]]$ is the weakest restriction  
       implying both $[[us'1]]$ and $[[us'2]]$,
       we have $[[Θ ⊢ us ⇒ us'1 & us'2]]$, as required.

@@

       This way, the algorithm terminates with the result $[[us0' \ {â⁺*}]]$
       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁺*} ]]$. Notice that
       $[[us0' \ {â⁺*} : Θ | uv uN' ]]$. This way, to show $[[ Θ ⊢ us ⇒ us0' \
       {â⁺*} ]]$. it suffices to consider an unification variable $[[β̂±]] ∊
       [[uv uN']]$ and show that $[[Γ, pbs ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
       reflexivity of weakening (\cref{lemma:weakening-reflexivity}) since
       $[[us0'(β̂±)]] = [[us(β̂±)]]$.
    \end{caseof}
\end{proof}

### 2023-05-18 proofs/tex/equiv-lemmas.tex
Add soundness of solution merge.

@@

>>>>

\begin{lemma}[Declarative equivalence is transitive]
  \hfill
  \label{lemma:decl-equiv-transitivity}
  \begin{itemize}
  \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
  \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{should be easy to do by induction since the types are getting smaller}
\end{proof}
\begin{lemma}[Algorithmization of declarative equivalence]
  \label{lemma:decl-equiv-algorithmization}
  Declarative equivalence is equality of normal forms. 
  \begin{itemize}
    \item[$+$] $[[iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
    \item[$-$] $[[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
  \end{itemize}
\end{lemma}
\begin{proof} \hfill
  \begin{itemize}
    \item[$+$] Let us prove both directions separately.
    \begin{itemize}
      \item[$\Rightarrow$] 
        exactly by \cref{lemma:normalization-completeness},
      \item[$\Leftarrow$] 
        from \cref{lemma:normalization-soundness}, we know
        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
        $[[iP ≈ iQ]]$.
    \end{itemize}
    \item[$-$] The proof is exactly the same.
  \end{itemize}
\end{proof}

@@

\begin{lemma}[Algorithmization of subtyping-induced equivalence]
  \label{lemma:subt-equiv-algorithmization}
  Mutual subtyping is equality of normal forms.
 Assuming all the types below are well-formed in $[[Γ]]$:
  \begin{itemize}
    \item [$+$] $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
    \item [$-$] $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  Let us prove the positive case, the negative case is symmetric.
  We prove both directions of $\iff$ separately:
  \begin{itemize}
    \item [$\Rightarrow$] exactly \cref{corollary:nf-complete-wrt-subt-equiv};
    \item [$\Leftarrow$] by \cref{lemma:decl-equiv-algorithmization,lemma:equiv-soundness}.
  \end{itemize}
\end{proof}
